---
title: "WHO Measles Outbreak Investigation"
format: 
  dashboard:
    orientation: rows
    logo:  "../images/calpolylogo.avif"
    nav-buttons:
      - icon: github
        href: https://github.com/jasminecabrera
        aria-label: GitHub
    embed-resources: true
theme: [united, styles.scss]
server: shiny
---

```{r setup}
#| message: false
#| context: setup

library(tidyverse)
library(ggridges)
library(ggrepel)
library(tidyverse)
library(car)
library(pscl)
library(gt)
library(broom)
library(MASS)
library(scales)
library(htmltools)
library(leaflet)
library(hexbin)
library(tigris)
library(sf)
library(rnaturalearth)
library(gganimate)
library(gifski)
library(png)
library(scales)
library(plotly)
library(shiny)

```

```{r}
#| context: data
measles <- read_csv(here::here("data", "measles_df_clean.csv"))
measles <- measles |>
  rename(`Nominal GDP per Capita` = `Nominal GDP`)

romania <- read_csv(here::here("data", "API_ROU_DS2_en_csv_v2_12273.csv"), 
                    skip = 3) |>
  filter(`Indicator Code` == "SE.PRM.CMPT.ZS") |>
  pivot_longer(cols = 5:last_col()) |>
  filter(!is.na(value)) |>
  mutate(name = as.numeric(name))

romania_tabset <- read_csv(here::here("data", "API_ROU_DS2_en_csv_v2_12273.csv"), 
                           skip = 3) |>
  filter(`Indicator Code` %in% c("SP.DYN.IMRT.IN",
                                 "NY.GDP.PCAP.PP.CD",
                                 "SI.POV.DDAY")) |>
  pivot_longer(cols = 5:last_col()) |>
  filter(!is.na(value)) |>
  mutate(name = as.numeric(name))

romania_school <- read_csv(here::here("data", "API_ROU_DS2_en_csv_v2_12273.csv"), 
                    skip = 3) |>
  filter(`Indicator Code` == "SE.PRM.ENRR") |>
  pivot_longer(cols = 5:last_col()) |>
  filter(!is.na(value)) |>
  mutate(name = as.numeric(name))

# get world polygons
world <- ne_countries(scale = "medium", returnclass = "sf")

# renaming Country names to match world dataset
measles_renamed <- measles |>
  filter(Year == 2024) |>
  mutate(
    Country = case_when(
      Country == "Congo" ~ "Republic of the Congo",
      Country == "Eswatini" ~ "Kingdom of eSwatini",
      Country == "Gambia" ~ "The Gambia",
      Country == "United Republic of Tanzania" ~ "Tanzania",
      Country == "Bolivia (Plurinational State of)" ~ "Bolivia",
      Country == "United States of America" ~ "United States",
      Country == "Venezuela (Bolivarian Republic of)" ~ "Venezuela",
      Country == "Iran (Islamic Republic of)" ~ "Iran",
      Country == "Syrian Arab Republic" ~ "Syria",
      Country == "Czechia" ~ "Czech Republic",
      Country == "Netherlands (Kingdom of the)" ~ "Netherlands",
      Country == "Republic of Moldova" ~ "Moldova",
      Country == "Türkiye" ~ "Turkey",
      Country == "United Kingdom of Great Britain and Northern Ireland" ~ "United Kingdom",
      Country == "Democratic People's Republic of Korea" ~ "Dem. Rep. Korea",
      Country == "Lao People's Democratic Republic" ~ "Lao PDR",
      Country == "Micronesia (Federated States of)" ~ "Federated States of Micronesia",
      Country == "Viet Nam" ~ "Vietnam",
      TRUE ~ Country
    )
  )

# join measles data to world polygons
world_measles <- world |>
  left_join(measles_renamed,
            by = c("name_long" = "Country"))
```
# Global Measles Data

## Research Question {width="100%", height="15%"}

::: {.card title="Research Question" color="#D4AF37"}

What socioeconomic factors have an effect on Measles infection rate? Are these socioeconomic factors correlated with vaccine coverage rate?

:::

## Socioeconomic Factors (Row 2) {height="85%"} 

### Heatmap

```{r}
#| title: Measles Cases per Million (M) Averages Across Time & Region Heatmap
#| alt-text: "The Americas do not have data for 2012 and 2013. The heatmap indicates that there was a measles outbreak from 2014-2016 in the Western Pacific region. Notably, the African countries seem to have relatively consistent measles cases over time, with a spike in 2019."

# create plot
measles_heatmap <- measles |>
  group_by(Region, Year) |>
  summarise(mean_cases = mean(`Total Measles per Million`, na.rm = TRUE),
            avg_pop = mean(`Total Pop.`, na.rm = TRUE),
            .groups = "drop") |>
  
  # compute log(1+x) since we 0s
  ggplot(aes(x = Year,
             y = Region,
             fill = log1p(mean_cases))) + 
  
  geom_tile(color = "white",
            aes(
            text = paste0(
      "Year: ", Year,
      "<br>Avg. Cases: ", format(round(mean_cases, 2), 
             big.mark = ",", 
             scientific = FALSE),
      "<br>Average Population: ", format(round(avg_pop, 2), big.mark = ",", scientific = F)
    ))) +
  
  # non-default colors
  scale_fill_continuous(palette = c("white", "red")) + 
  guides(fill = guide_colorbar(direction = "horizontal")) +
  
  # non-default theme
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 12),
        legend.position = "top") +
  
  # labels 
  labs(x = "Year",
       y = "",
       fill = "Log(M + 1)",
       caption = "Unlike other regions, the Americas do not have data for 2012 and 2013.") +
  scale_x_continuous(breaks = seq(min(measles$Year),
                                  max(measles$Year),
                                  by = 2))

ggplotly(measles_heatmap, tooltip = "text") |>
  layout(legend = list(orientation = "h", x = 0.5, xanchor = "center", y = 1.1))
```

#### MCV1 Coverage {.tabset}

```{r}
#| title: Average MCV1 Coverage
label_data <- measles |>
  group_by(Region, Year) |>
  summarise(mean_coverage = mean(`MCV1 Coverage`, na.rm = TRUE),
            .groups = "drop") |>
  drop_na() |>
  group_by(Region) |>
  slice_max(Year) |>
  arrange(mean_coverage)

# make plot
mcv1_plot <- measles |>
  group_by(Region, Year) |>
  summarise(mean_coverage = mean(`MCV1 Coverage`, na.rm = TRUE), 
            .groups = "drop") |>
  
  ggplot(aes(x = Year, 
             y = mean_coverage, 
             color = Region)) +
  geom_line(linewidth = 1) +
  geom_point() +
  
  # labels
  labs(
       y = "",
       x = "Year") +
  
  # unique colors
  scale_color_manual(values = c("Africa" = "#d35e7f",
                                  "Americas" = "#9fc7b9",
                                  "Eastern Mediterranean" = "#a999eb",
                                  "Europe" = "#9bcdef",
                                  "South East Asia" = "#ff9d60",
                                  "Western Pacific" = "#d4a9ce")) +
  scale_fill_manual(values = c("Africa" = "#d35e7f",
                                  "Americas" = "#9fc7b9",
                                  "Eastern Mediterranean" = "#a999eb",
                                  "Europe" = "#9bcdef",
                                  "South East Asia" = "#ff9d60",
                                  "Western Pacific" = "#d4a9ce")) +
  
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  
  # edit graph  
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"),
        legend.position = "none") +
  
  # move labels
  geom_label_repel(data = label_data,
                  aes(x = Year, y = mean_coverage, 
                      label = Region, fill = Region),
                      inherit.aes = FALSE,
                      nudge_x = 0.5,
                      direction = "y",
                      hjust = 0,
                      segment.color = NA) +
  
  # expand grid
  expand_limits(x = max(label_data$Year) + 5)

mcv1_plot
```

```{r}
#| title: Demographics Correlations
measles |>
  filter(!is.na(`MCV1 Coverage`),
         !is.na(`Nominal GDP per Capita`),
         !is.na(`Total Measles per Million`)) |>
  split(~ Region) |>
  map_dfr(function(x) {
    tibble(Region = first(x$Region),
           `MCV1 vs GDP` = cor(x$`MCV1 Coverage`,
                                     x$`Nominal GDP per Capita`,
                                     use = "complete.obs"),
           `MCV1 vs Measles` = cor(x$`MCV1 Coverage`,
                                         x$`Nominal GDP per Capita`,
                                         use = "complete.obs"),
           `GDP vs Measles` = cor(x$`Nominal GDP per Capita`,
                                        x$`Total Measles per Million`,
                                        use = "complete.obs"))}) |>
  arrange(Region) |>
  gt() |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  
  # format values
  fmt_number(columns = c(`MCV1 vs GDP`, `MCV1 vs Measles`, `GDP vs Measles`),
    decimals = 3,
    use_seps = TRUE)
```

```{r}
#| title: MCV1 vs. GDP Plot
measles |>
  filter(!is.na(`Nominal GDP per Capita`),
         `Nominal GDP per Capita` > 0,
         !is.na(`MCV1 Coverage`)) |>
  ggplot(aes(x = log(`Nominal GDP per Capita`),
             y = `MCV1 Coverage`,
             fill = Region)) + 
  geom_hex() + 
  facet_wrap(~Region) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  scale_fill_manual(values = c(
    "Africa" = "#d35e7f",
    "Americas" = "#9fc7b9",
    "Eastern Mediterranean" = "#a999eb",
    "Europe" = "#9bcdef",
    "South East Asia" = "#ff9d60",
    "Western Pacific" = "#d4a9ce")) +
  labs(x = "Log Nominal GDP per Capita (USD)",
       y = NULL,
       subtitle = "MCV1 Coverage vs. Log Nominal GDP per Capita") +
  theme_bw() +
  theme(legend.position = "none") +
  scale_x_continuous(labels = \(x) paste0("$", x)) 
```


### Num Cases Leaflet Plot

```{r}
#| title: Global Measles Cases in 2024

# same colors as MCV1 plot
region_colors <- c(
  "Africa" = "#d35e7f",
  "Americas" = "#9fc7b9",
  "Eastern Mediterranean" = "#a999eb",
  "Europe" = "#9bcdef",
  "South-East Asia" = "#ff9d60",
  "Western Pacific" = "#d4a9ce"
)

# assign colors based on Region
pal <- colorFactor(
  palette = region_colors,
  domain = world_measles$Region,
  na.color = "transparent"
)

# leaflet map
leaflet(world_measles) |>
  addProviderTiles("CartoDB.Positron") |>
  
  addPolygons(
    fillColor = ~pal(Region),
    fillOpacity = 0.7,
    
    color = "black",
    weight = 0.5,
    
    popup = ~paste0(
      "<strong>", name, "</strong><br>",
      "Total Measles per Million: ",
      `Total Measles per Million`, "<br>",
      "MCV1 Coverage: ",
      `MCV1 Coverage`, "% <br>",
      "Nominal GDP per Capita: $",
      format(`Nominal GDP per Capita`, 
             big.mark = ",", 
             scientific = FALSE)
    ),
    
    label = ~name,
    
    highlightOptions = highlightOptions(
      weight = 2,
      color = "white",
      bringToFront = TRUE
    )
  ) |>
  
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~Region[!is.na(Region)],
    title = "Region"
  )
```





# Romania Measles Data

## Romania: representative case {height="20%}
::: {.card title="Why Romania?"}

Based on our Global Annual Measles heatmap, we observed that the Western Pacific region had the highest measles cases per million from 2014-2016. However, most recent data from 2024 shows that Europe now leads in measles cases per million. Focusing on Europe, we
found that Romania is an informative case: its population of 19.07 million is 
roughly representative of the region, and it reported the largest total measles
cases and the third highest average cases in Europe 2024, making it an interest
for further analysis.
:::

## Second Row: Animated Plot {height="60%"}

### Animated Plot

```{r}
#| title: Romania Measles Cases Trend Compared to the Rest of the World
measles |>
  mutate(Country = if_else(Country == "Romania", "Romania", "Rest of World")) |>
  group_by(Country, Year) |>
  summarize(mean_cases = mean(`Total Measles per Million`, 
                              na.rm = TRUE)) |>
  drop_na() |>
  ungroup() |>
  ggplot(aes(x = Year, y = mean_cases)) +
  geom_line(color = "orange") +
  geom_point(color = "darkorange") + 
    labs(x = "Year",
         y = NULL,
         subtitle = "Average Total Measles Cases per Million") +
    theme_minimal() + 
  theme(panel.grid.minor = element_blank(),
        legend.position = "none") +
  transition_states(Country, transition_length = 2, state_length = 1) + 
  ease_aes("cubic-in-out") + 
  ggtitle("Measles Case Trend for {closest_state}")

```

### Reactive Output
```{r}
#| content: card-toolbar
#| title: "Socioeconomic Data"


selectInput('data_type', 'Select Data', 
            c("Education & Health", "Infant Mortality Rate", "GDP"),
            width = "200px")
```

```{r}
#| context: server
#| message: false
output$RomaniaData <- renderPlot({
  
  if (input$data_type == "Education & Health") {
    
    p <- measles |>
  filter(Country == "Romania") |>
  filter(!is.na(`MCV1 Coverage`)) |>
  left_join(romania, by = join_by("Year" == "name")) |>
  left_join(romania_school, by = join_by("Year" == "name")) |>
  rename(
    "Primary School Enrollment" = value.y,
    "Primary School Completion Rate" = value.x,
    "MCV1 Coverage" = `MCV1 Coverage`
  ) |>
  pivot_longer(
    cols = c(`MCV1 Coverage`, 
             `Primary School Enrollment`,
             `Primary School Completion Rate`), 
    names_to = "series",
    values_to = "value"
  ) |>
  mutate(series = factor(series, levels = c(
    "MCV1 Coverage",
    "Primary School Enrollment",
    "Primary School Completion Rate"
  ))) |>
  ggplot(aes(x = Year, y = value, color = series)) + 
  geom_line() + 
  geom_point() +
  labs(title = "Romania: Education", x = "Year", y = "",
       color = "Data") +
  theme_bw() +
  theme(legend.position = "top") +
  scale_y_continuous(labels = \(x) paste0(x, "%"))
    
  } else if (input$data_type == "Infant Mortality Rate") {
    
    p <- romania_tabset |>
      filter(`Indicator Code` == "SP.DYN.IMRT.IN") |>
      ggplot(aes(x = name, y = value)) +
      geom_line(color = "orange") + geom_point(color = "darkorange") +
      xlim(c(2014, 2024)) + 
      labs(title = "Infant Mortality Rate", x = "Year", y = "") +
      theme_bw() +
      scale_y_continuous(labels = \(x) paste0(x, "%"))
    
  } else {
    
    p <- romania_tabset |>
      filter(`Indicator Code` == "NY.GDP.PCAP.PP.CD") |>
      ggplot(aes(x = name, y = value)) +
      geom_line(color = "orange") + 
      geom_point(color = "darkorange") +
      xlim(c(2014, 2024)) +
      labs(title = "GDP per Capita", x = "Year", y = "") +
      theme_bw() +
      scale_y_continuous(labels = \(x) paste0("$", comma(x)))
  }
  
  p
  
})
```

```{r}
plotOutput('RomaniaData')
```

## Third Row: Analysis {height="20%"}
:::{.card title="Findings"}
Examining potential contributing factors, we note that while measles vaccination
coverage (MCV1) has decreased, other socioeconomic indicators such as GDP and
poverty level have remained relatively stable. Interestingly, primary school enrollment
and completion rates show a similar declining trend to MCV1 coverage, suggesting a 
possible link between education levels, vaccination uptake, and the observed 
increase in measles cases.
:::






