---
title: "WHO Measles Outbreak Investigation"
format: 
  dashboard:
    orientation: columns
    logo:  "calpolylogo.avif"
    nav-buttons:
      - icon: github
        href: https://github.com/rjimenezdata
        aria-label: GitHub
    embed-resources: true
theme: united
---

```{r setup}
#| message: false

library(tidyverse)
library(ggridges)
library(ggrepel)
library(tidyverse)
library(car)
library(pscl)
library(gt)
library(broom)
library(MASS)
library(scales)
library(htmltools)
library(leaflet)
library(hexbin)
library(tigris)
library(sf)
library(rnaturalearth)
library(gganimate)
library(gifski)
library(png)
library(scales)

measles <- read_csv(here::here("data", "measles_df_clean.csv"))
measles <- measles |>
  rename(`Nominal GDP per Capita` = `Nominal GDP`)

romania <- read_csv(here::here("data", "API_ROU_DS2_en_csv_v2_12273.csv"), 
                    skip = 3) |>
  filter(`Indicator Code` == "SE.PRM.CMPT.ZS") |>
  pivot_longer(cols = 5:last_col()) |>
  filter(!is.na(value)) |>
  mutate(name = as.numeric(name))

romania_tabset <- read_csv(here::here("data", "API_ROU_DS2_en_csv_v2_12273.csv"), 
                           skip = 3) |>
  filter(`Indicator Code` %in% c("SP.DYN.IMRT.IN",
                                 "NY.GDP.PCAP.PP.CD",
                                 "SI.POV.DDAY")) |>
  pivot_longer(cols = 5:last_col()) |>
  filter(!is.na(value)) |>
  mutate(name = as.numeric(name))

romania_school <- read_csv(here::here("data", "API_ROU_DS2_en_csv_v2_12273.csv"), 
                    skip = 3) |>
  filter(`Indicator Code` == "SE.PRM.ENRR") |>
  pivot_longer(cols = 5:last_col()) |>
  filter(!is.na(value)) |>
  mutate(name = as.numeric(name))
```

# Global Measles Data

## Annual Measles Cases Data (Left Column) {width="40%"}

### Annual Measles Cases

```{r}
#| title: Global Annual Measles Cases Rates per Million

annual_measles_tbl <- measles |>
  filter(!is.na(`Total Measles per Million`),
    !is.na(`Total Pop.`)) |>
  split(~ Year + Region) |>
  map_dfr(function(x) 
    {tibble(Year = first(x$Year),
     Region = first(x$Region),
     `Avg. Measles per Million` = mean(x$`Total Measles per Million`, na.rm = TRUE),
     `Avg. Pop.` = mean(x$`Total Pop.`, na.rm = TRUE))}) |>
  arrange(desc(`Avg. Measles per Million`)) |>
  filter(!is.na(Year)) |>
  gt() |>
  tab_footnote(footnote = "Data is arranged is descending order by Average Measles Cases per Million.") |>
  
  # format numbers
  fmt_number(columns = c(`Avg. Pop.`, `Avg. Measles per Million`), use_seps = T) |>
  fmt_number(columns = `Avg. Pop.`, decimals = 0) |>
  
  # labels
  cols_label(Year = md("**Year**"),
             Region = md("**Region**"),
             "Avg. Pop." = md("**Avg. Pop.**"),
             "Avg. Measles per Million" = md("**Avg. Measles per Million**")) |>
  
  # add heat maps
  data_color(columns = `Avg. Measles per Million`,
             fn = col_numeric(palette = c("white", "red"),
                              domain = NULL))

# wrap in scrollable div
browsable(div(style = "max-height:500px; overflow-y:auto;", annual_measles_tbl))
```


### Num Cases Leaflet Plot

```{r}
#| title: Global Measles Cases in 2024

# icon
measles_icon <- makeIcon(
  iconUrl = "https://static.vecteezy.com/system/resources/thumbnails/017/177/954/small/round-medical-cross-symbol-on-transparent-background-free-png.png",
  iconWidth = 10,
  iconHeight = 10)

# leaflet map
leaflet(data = measles |> filter(Year == 2024)) |>
  addProviderTiles("CartoDB.Positron") |>
  
  addMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    
    icon = measles_icon,
    
    # Total Measles per Million appears when clicked, not just hovered over
    popup = ~paste0(
      "<strong>", Country, " (", Year, ")", "</strong><br>",
      "<strong>", "Region: ", Region, "</strong><br>",
      "Total Measles per Million: ",
      `Total Measles per Million`, "<br>",
      "MCV1 Coverage: ",
      `MCV1 Coverage`, "% <br>",
      "Nominal GDP per Capita: $",
      format(`Nominal GDP per Capita`, 
             big.mark = ",", 
             scientific = FALSE)),
    
    # country name appears when hovered over
    label = ~paste(Country))
  
```

## Socioeconomic Factors (Right Column) {width="60%"}

### Heatmap

```{r}
#| title: Average Measles Cases Rates per Million Heatmap
#| alt-text: "The Americas do not have data for 2012 and 2013. The heatmap indicates that there was a measles outbreak from 2014-2016 in the Western Pacific region. Notably, the African countries seem to have relatively consistent measles cases over time, with a spike in 2019."

# create plot
measles_heatmap <- measles |>
  group_by(Region, Year) |>
  summarise(mean_cases = mean(`Total Measles per Million`, na.rm = TRUE),
            .groups = "drop") |>
  
  # compute log(1+x) since we 0s
  ggplot(aes(x = Year,
             y = Region,
             fill = log1p(mean_cases))) + 
  geom_tile(color = "white") +
  
  # non-default colors
  scale_fill_continuous(palette = c("white", "red")) + 
  guides(fill = guide_colorbar(direction = "horizontal")) +
  
  # non-default theme
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.title = element_text(size = 12),
        legend.position = "top") +
  
  # labels and alt text
  labs(x = "Year",
       y = "",
       fill = "Log (Measles Cases Per Million +1)",
       caption = "Unlike other regions, the Americas do not have data for 2012 and 2013.") +
  scale_x_continuous(breaks = seq(min(measles$Year),
                                  max(measles$Year),
                                  by = 2))
measles_heatmap
```

### MCV1 Coverage {.tabset}

```{r}
#| title: Average MCV1 Coverage for 1 Year Olds Across Time and Region (WHO)
label_data <- measles |>
  group_by(Region, Year) |>
  summarise(mean_coverage = mean(`MCV1 Coverage`, na.rm = TRUE),
            .groups = "drop") |>
  drop_na() |>
  group_by(Region) |>
  slice_max(Year) |>
  arrange(mean_coverage)

# make plot
mcv1_plot <- measles |>
  group_by(Region, Year) |>
  summarise(mean_coverage = mean(`MCV1 Coverage`, na.rm = TRUE), 
            .groups = "drop") |>
  
  ggplot(aes(x = Year, 
             y = mean_coverage, 
             color = Region)) +
  geom_line(linewidth = 1) +
  geom_point() +
  
  # labels
  labs(
       y = "",
       x = "Year") +
  
  # unique colors
  scale_color_manual(values = c("Africa" = "#d35e7f",
                                  "Americas" = "#9fc7b9",
                                  "Eastern Mediterranean" = "#a999eb",
                                  "Europe" = "#9bcdef",
                                  "South East Asia" = "#ff9d60",
                                  "Western Pacific" = "#d4a9ce")) +
  scale_fill_manual(values = c("Africa" = "#d35e7f",
                                  "Americas" = "#9fc7b9",
                                  "Eastern Mediterranean" = "#a999eb",
                                  "Europe" = "#9bcdef",
                                  "South East Asia" = "#ff9d60",
                                  "Western Pacific" = "#d4a9ce")) +
  
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  
  # edit graph  
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"),
        legend.position = "none") +
  
  # move labels
  geom_label_repel(data = label_data,
                  aes(x = Year, y = mean_coverage, 
                      label = Region, fill = Region),
                      inherit.aes = FALSE,
                      nudge_x = 0.5,
                      direction = "y",
                      hjust = 0,
                      segment.color = NA) +
  
  # expand grid
  expand_limits(x = max(label_data$Year) + 5)

mcv1_plot
```

```{r}
#| title: Correlations Between Demographic Variables
measles |>
  filter(!is.na(`MCV1 Coverage`),
         !is.na(`Nominal GDP per Capita`),
         !is.na(`Total Measles per Million`)) |>
  split(~ Region) |>
  map_dfr(function(x) {
    tibble(Region = first(x$Region),
           `MCV1 vs GDP` = cor(x$`MCV1 Coverage`,
                                     x$`Nominal GDP per Capita`,
                                     use = "complete.obs"),
           `MCV1 vs Measles` = cor(x$`MCV1 Coverage`,
                                         x$`Nominal GDP per Capita`,
                                         use = "complete.obs"),
           `GDP vs Measles` = cor(x$`Nominal GDP per Capita`,
                                        x$`Total Measles per Million`,
                                        use = "complete.obs"))}) |>
  arrange(Region) |>
  gt() |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  
  # format values
  fmt_number(columns = c(`MCV1 vs GDP`, `MCV1 vs Measles`, `GDP vs Measles`),
    decimals = 3,
    use_seps = TRUE)
```

```{r}
#| title: MCV1 Coverage vs. GDP Plot - Regional
measles |>
  filter(!is.na(`Nominal GDP per Capita`),
         `Nominal GDP per Capita` > 0,
         !is.na(`MCV1 Coverage`)) |>
  ggplot(aes(x = log(`Nominal GDP per Capita`),
             y = `MCV1 Coverage`,
             fill = Region)) + 
  geom_hex() + 
  facet_wrap(~Region) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  scale_fill_manual(values = c(
    "Africa" = "#d35e7f",
    "Americas" = "#9fc7b9",
    "Eastern Mediterranean" = "#a999eb",
    "Europe" = "#9bcdef",
    "South East Asia" = "#ff9d60",
    "Western Pacific" = "#d4a9ce")) +
  labs(x = "Log Nominal GDP per Capita (USD)",
       y = "MCV1 Coverage") +
  theme_bw() +
  theme(legend.position = "none") 
```

# Romania Measles Data

## Left Column

### Romania: representative case {height=35%}
::: {.card title="Motivations for Romania, a representative case."}

Based on our Global Annual Measles Cases Rates per Million table and accompanying
heatmap, we observed that the Western Pacific region had the highest measles 
cases per million between 2014 and 2016. However, the most recent data from 2024
shows that Europe now leads in measles cases per million. Focusing on Europe, we
found that Romania is an informative case: its population of 19.07 million is 
roughly representative of the region, and it reported the largest total measles
cases and the third highest average cases in Europe 2024, making it an interest
for further analysis.
:::

### Average MCV1 Coverage Compared to {height=65%}

```{r}
#| title: Average MCV1 Coverage Compared to Primary School Enrollment & Completion Rate
# make plot

measles |>
  filter(Country == "Romania") |>
  filter(!is.na(`MCV1 Coverage`)) |>
  left_join(romania, by = join_by("Year" == "name")) |>
  left_join(romania_school, by = join_by("Year" == "name")) |>
  rename("Primary School Enrollment" = value.y,
         "Primary School Completion Rate (%)" = value.x,
         "MCV1 Coverage (%)" = `MCV1 Coverage`) |>
  pivot_longer(cols = c(`MCV1 Coverage (%)`, 
                        `Primary School Enrollment`,
                        `Primary School Completion Rate (%)`), 
               names_to = "series",
    values_to = "value") |>
  mutate(series = factor(series, levels = c(
  "MCV1 Coverage (%)",
  "Primary School Enrollment",
  "Primary School Completion Rate (%)"))) |>
  ggplot(aes(x = Year, y = value)) + 
  geom_line(color = "orange") + 
  geom_point(color = "darkorange") +
  labs(title = "Romania: {closest_state}",
       x = "Year",
       y = "",
       color = "") +
  theme_minimal() + 
  theme(panel.grid.minor = element_blank()) + 
  scale_x_continuous(breaks = c(2015, 2020, 2024)) + 
  scale_y_continuous(labels = scales::percent_format(scale = 1), 
                     limits = c(70, 100)) + 
  transition_states(series, 
                    transition_length = 2,
                    state_length = 1)
```

## Right Column 
### Row height {height="68%"}
### Data {.tabset} 
```{r}
#| title: Infant Mortality Rate
romania_tabset |>
  filter(`Indicator Code` == "SP.DYN.IMRT.IN") |>
  ggplot(aes(x = name, y = value)) +
  geom_line(color = "orange") +
  geom_point(color = "darkorange") +
  xlim(c(2014, 2024)) +
  ylim(c(0, 15)) +
  labs(x = "Year",
       y = "",
       title = "Infant Mortality Rate (per 1,000 live births)") +
  theme_bw()
```
```{r}
#| title: GDP per Capita
romania_tabset |>
  filter(`Indicator Code` == "NY.GDP.PCAP.PP.CD") |>
  ggplot(aes(x = name, y = value)) +
  geom_line(color = "orange") +
  geom_point(color = "darkorange") +
  xlim(c(2014, 2024)) +
  labs(x = "Year",
       y = "",
       title = "GDP per Capita ($)") +
  theme_bw()
```
```{r}
#| title: Poverty Headcount at $3.00 a day
romania_tabset |>
  filter(`Indicator Code` == "SI.POV.DDAY") |>
  ggplot(aes(x = name, y = value)) +
  geom_line(color = "orange") +
  geom_point(color = "darkorange") +
  xlim(c(2014, 2024)) +
  labs(x = "Year",
       y = "",
       title = "Percentage of the Population Living on Less than $3.00 a Day") +
  theme_bw()
```


### Analysis {height="32%"}
:::{.card title="Exploratory Findings"}
Examining potential contributing factors, we note that while measles vaccination
coverage (MCV1) has decreased, other socioeconomic indicators such as GDP and
poverty level have remained relatively stable. Interestingly, primary school enrollment
and completion rates show a similar declining trend to MCV1 coverage, suggesting a 
possible link between education levels, vaccination uptake, and the observed 
increase in measles cases.
:::
