---
title: "WHO Measles Outbreak Investigation"
format: 
  dashboard:
    orientation: columns
    logo:  "../images/calpolylogo.avif"
    nav-buttons:
      - icon: github
        href: https://github.com/rjimenezdata
        aria-label: GitHub
    embed-resources: true
theme: united
---

```{r setup}
#| message: false

library(tidyverse)
library(ggridges)
library(ggrepel)
library(tidyverse)
library(car)
library(pscl)
library(gt)
library(broom)
library(MASS)
library(scales)
library(htmltools)
library(leaflet)
library(hexbin)
library(tigris)
library(sf)
library(rnaturalearth)
library(gganimate)
library(gifski)
library(png)
library(scales)

measles <- read_csv(here::here("data", "measles_df_clean.csv"))
measles <- measles |>
  rename(`Nominal GDP per Capita` = `Nominal GDP`)

romania <- read_csv(here::here("data", "API_ROU_DS2_en_csv_v2_12273.csv"), skip = 3) |>
  filter(`Indicator Code` == "SE.PRM.CMPT.ZS") |>
  pivot_longer(cols = 5:last_col()) |>
  filter(!is.na(value)) |>
  mutate(name = as.numeric(name))
```

# Global Measles Data

## Annual Measles Cases Data

### Annual Measles Cases {.tabset}

```{r}
#| title: Global Annual Measles Cases Rates per Million

annual_measles_tbl <- measles |>
  filter(!is.na(`Total Measles per Million`),
    !is.na(`Total Pop.`)) |>
  split(~ Year + Region) |>
  map_dfr(function(x) 
    {tibble(Year = first(x$Year),
     Region = first(x$Region),
     `Avg. Measles per Million` = mean(x$`Total Measles per Million`, na.rm = TRUE),
     `Avg. Pop.` = mean(x$`Total Pop.`, na.rm = TRUE))}) |>
  arrange(desc(`Avg. Measles per Million`)) |>
  filter(!is.na(Year)) |>
  gt() |>
  tab_footnote(footnote = "Data is arranged is descending order by Average Measles Cases per Million.") |>
  
  # format numbers
  fmt_number(columns = c(`Avg. Pop.`, `Avg. Measles per Million`), use_seps = T) |>
  fmt_number(columns = `Avg. Pop.`, decimals = 0) |>
  
  # labels
  cols_label(Year = md("**Year**"),
             Region = md("**Region**"),
             "Avg. Pop." = md("**Avg. Pop.**"),
             "Avg. Measles per Million" = md("**Avg. Measles per Million**")) |>
  
  # add heat maps
  data_color(columns = `Avg. Measles per Million`,
             fn = col_numeric(palette = c("white", "red"),
                              domain = NULL))

# wrap in scrollable div
browsable(div(style = "max-height:500px; overflow-y:auto;", annual_measles_tbl))
```

```{r}
#| title: Annual Change in Measles Cases Rates per Million - Western Pacific Region
yoy_change <- function(df, region_name) {
  
  # input checks
  required_cols <- c("Year", "Region", "Total Measles per Million")
  
  if (!all(required_cols %in% names(df))) {
    missing_cols <- required_cols[!required_cols %in% names(df)]
    stop(paste("Missing required columns:", paste(missing_cols, collapse = ", ")))}
  
  if (!is.character(region_name) || length(region_name) != 1) {
    stop("Country Name must be a valid country.")}
  
  # case when region is not in dataset (handles capitalization)
  if (!tolower(region_name) %in% c("south east asia", "europe", 
                                    "western pacific", "americas", 
                                    "eastern mediterranean", "africa")){
    stop("Invalid Region Name. Please check spelling. Options are South East Asia, Europe,
         Western Pacific, America, Eastern Mediterranean, and Africa.")}
  
  # make sure first letter is capitalized
  if (tolower(region_name) %in% c("south east asia", "europe", 
                                    "western pacific", "americas", 
                                    "eastern mediterranean", "africa")){
    region_name <- stringr::str_to_title(region_name)}
  
  # empty dataset
  if (nrow(df) == 0) {
    warning("No data found.")}

  # create summary table
    df |>
    filter(!is.na(`Total Measles per Million`),
           !is.na(`Total Pop.`)) |>
    group_by(Year, Region) |>
    summarize(`Avg. Measles per Million` = mean(`Total Measles per Million`, na.rm = T),
              `Avg. Pop.` = mean(`Total Pop.`, na.rm = T),
              .groups = "drop") |>
    filter(Region == region_name) |>
    arrange(Year) |> 
    mutate(`Measles Rate Change` = `Avg. Measles per Million` - 
             lag(`Avg. Measles per Million`)) |>
    dplyr::select(Year, `Avg. Pop.`, `Avg. Measles per Million`, 
           `Measles Rate Change`) |>
    gt() |>
      
    # format numbers
    fmt_number(columns = c(`Avg. Pop.`, `Avg. Measles per Million`, 
                         `Measles Rate Change`), use_seps = T) |>
    fmt_number(columns = `Avg. Pop.`, decimals = 0) |>
      
    # labels
    cols_label(Year = md("**Year**"),
               "Avg. Pop." = md("**Avg. Pop.**"),
               "Avg. Measles per Million" = md("**Avg. Measles per Million**"),
               "Measles Rate Change" = md("**Measles Rate Change**"))} |>
  
  # color
  data_color(columns = `Measles Rate Change`,
             fn = col_numeric(
               palette = c("white", "red"),
               domain = NULL))

yoy_change(measles, "Western Pacific")
```

### Num Cases Leaflet Plot


```{r}
#| title: Cases in 2024 vs MCV1 Coverage vs. GDP per Capita

# icon
measles_icon <- makeIcon(
  iconUrl = "https://cdn-icons-png.flaticon.com/512/13297/13297903.png",
  iconWidth = 10,
  iconHeight = 10
)

# leaflet map
leaflet(data = measles |> filter(Year == 2024)) |>
  addProviderTiles("CartoDB.Positron") |>
  
  addMarkers(
    lng = ~Longitude,
    lat = ~Latitude,
    
    icon = measles_icon,
    
    # Total Measles per Million appears when clicked, not just hovered over
    popup = ~paste0(
      "<strong>", Country, "</strong><br>",
      "<strong>", Year, "</strong><br>",
      "Total Measles per Million: ",
      `Total Measles per Million`, "<br>",
      "MCV1 Coverage: ",
      `MCV1 Coverage`, "% <br>",
      "GDP per Capita: $",
      `Nominal GDP per Capita`
      
    ),
    
    # country name appears when hovered over
    label = ~paste0(
      Country
    )
  )
  
```

## Socioeconomic Factors

### MCV1 Coverage

```{r}
#| title: Average MCV1 Coverage for 1 Year Olds Across Time and Region (WHO)
label_data <- measles |>
  group_by(Region, Year) |>
  summarise(mean_coverage = mean(`MCV1 Coverage`, na.rm = TRUE),
            .groups = "drop") |>
  drop_na() |>
  group_by(Region) |>
  slice_max(Year) |>
  arrange(mean_coverage)

# make plot
mcv1_plot <- measles |>
  group_by(Region, Year) |>
  summarise(mean_coverage = mean(`MCV1 Coverage`, na.rm = TRUE), 
            .groups = "drop") |>
  
  ggplot(aes(x = Year, 
             y = mean_coverage, 
             color = Region)) +
  geom_line(linewidth = 1) +
  geom_point() +
  
  # labels
  labs(
       y = "",
       x = "Year") +
  
  # unique colors
  scale_color_manual(values = c("Africa" = "#d35e7f",
                                  "Americas" = "#9fc7b9",
                                  "Eastern Mediterranean" = "#a999eb",
                                  "Europe" = "#9bcdef",
                                  "South East Asia" = "#ff9d60",
                                  "Western Pacific" = "#d4a9ce")) +
  scale_fill_manual(values = c("Africa" = "#d35e7f",
                                  "Americas" = "#9fc7b9",
                                  "Eastern Mediterranean" = "#a999eb",
                                  "Europe" = "#9bcdef",
                                  "South East Asia" = "#ff9d60",
                                  "Western Pacific" = "#d4a9ce")) +
  
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  
  # edit graph  
  theme_bw() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        axis.title = element_text(face = "bold"),
        legend.position = "none") +
  
  # move labels
  geom_label_repel(data = label_data,
                  aes(x = Year, y = mean_coverage, 
                      label = Region, fill = Region),
                      inherit.aes = FALSE,
                      nudge_x = 0.5,
                      direction = "y",
                      hjust = 0,
                      segment.color = NA) +
  
  # expand grid
  expand_limits(x = max(label_data$Year) + 5)

mcv1_plot
```

### Tables{.tabset}

```{r}
#| title: Correlations Between Demographic Variables
measles |>
  filter(!is.na(`MCV1 Coverage`),
         !is.na(`Nominal GDP per Capita`),
         !is.na(`Total Measles per Million`)) |>
  split(~ Region) |>
  map_dfr(function(x) {
    tibble(Region = first(x$Region),
           `MCV1 vs GDP` = cor(x$`MCV1 Coverage`,
                                     x$`Nominal GDP per Capita`,
                                     use = "complete.obs"),
           `MCV1 vs Measles` = cor(x$`MCV1 Coverage`,
                                         x$`Nominal GDP per Capita`,
                                         use = "complete.obs"),
           `GDP vs Measles` = cor(x$`Nominal GDP per Capita`,
                                        x$`Total Measles per Million`,
                                        use = "complete.obs"))}) |>
  arrange(Region) |>
  gt() |>
  tab_style(style = cell_text(weight = "bold"),
            locations = cells_column_labels()) |>
  
  # format values
  fmt_number(columns = c(`MCV1 vs GDP`, `MCV1 vs Measles`, `GDP vs Measles`),
    decimals = 3,
    use_seps = TRUE)
```

```{r}
#| title: MCV1 Coverage vs. GDP Plot - Regional
measles |>
  filter(!is.na(`Nominal GDP per Capita`),
         `Nominal GDP per Capita` > 0,
         !is.na(`MCV1 Coverage`)) |>
  ggplot(aes(x = log(`Nominal GDP per Capita`),
             y = `MCV1 Coverage`,
             fill = Region)) + 
  geom_hex() + 
  facet_wrap(~Region) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  scale_fill_manual(values = c(
    "Africa" = "#d35e7f",
    "Americas" = "#9fc7b9",
    "Eastern Mediterranean" = "#a999eb",
    "Europe" = "#9bcdef",
    "South East Asia" = "#ff9d60",
    "Western Pacific" = "#d4a9ce")) +
  labs(x = "Log Nominal GDP per Capita (USD)",
       y = "MCV1 Coverage") +
  theme_bw() +
  theme(legend.position = "none") 
```

```{r}
#| title: MCV1 Coverage vs. GDP Table - Western Pacific Countries
mcv_gdp <- function(region = NULL, country = NULL, year = NULL){
  
  # input checks
    # check that at least region or country is provided
  if (is.null(region) & is.null(country)) {
    stop("Please provide at least a region or a country.")}
  
  # valid regions
  valid_regions <- c("South East Asia", "Europe", 
                     "Western Pacific", "Americas", 
                     "Eastern Mediterranean", "Africa")
  
  # check region only if provided
  if (!is.null(region)) {
    if (!is.character(region) || length(region) != 1) {
      stop("Region must be a single character string.")}
    
    if (!tolower(region) %in% tolower(valid_regions)) {
      stop("Invalid region. Options are: South East Asia, Europe, Western Pacific, Americas, Eastern Mediterranean, Africa.")}
    region <- str_to_title(region)}
  
  # check country only if provided
  if (!is.null(country)) {
    if (!is.character(country) || length(country) != 1) {
      stop("Country must be a single character string.")}
    
    if (!tolower(country) %in% tolower(measles$Country)) {
      stop("Country not found in dataset. Please check spelling.")}}
  
  # check year only if provided
  if (!is.null(year)) {
    if (!is.numeric(year) || length(year) != 1 || year %% 1 != 0) {
      stop("Year must be a single integer.")
    }
    if (!year %in% measles$Year) {
      warning("Year not found in dataset.")}}
  
  # given country and year
  if (!is.null(country) & !is.null(year)) {
    browsable(div(style = "max-height:500px; overflow-y:auto;", 
      measles |>
      dplyr::filter(Country == country, Year == year) |>
      dplyr::select(Country, `Nominal GDP per Capita`, 
                    `MCV1 Coverage`, `Total Measles per Million`) |>
        arrange(desc(`Total Measles per Million`)) |>
        rename("Nominal GDP per Capita ($ USD)" = `Nominal GDP per Capita`) |>
      gt() |>
      fmt_number(columns = c(`Nominal GDP per Capita ($ USD)`), decimals = 3) |>
      fmt(columns = `MCV1 Coverage`,
          fns = function(x) paste0(x, "%")) |>
      tab_style(style = cell_text(weight = "bold"),
                locations = cells_column_labels(everything()))))
    
  # given region and year
  } else if (!is.null(region) & !is.null(year)) {
    browsable(div(style = "max-height:500px; overflow-y:auto;", 
      measles |>
      dplyr::filter(Region == region, Year == year) |>
      dplyr::select(Country, `Nominal GDP per Capita`, 
                    `MCV1 Coverage`, `Total Measles per Million`) |>
        arrange(desc(`Total Measles per Million`)) |>
        rename("Nominal GDP per Capita ($ USD)" = `Nominal GDP per Capita`) |>
      gt() |>
      fmt_number(columns = c(`Nominal GDP per Capita ($ USD)`), decimals = 3) |>
      fmt(columns = `MCV1 Coverage`,
          fns = function(x) paste0(x, "%")) |>
      tab_style(style = cell_text(weight = "bold"),
                locations = cells_column_labels(everything()))))
    
  # given only country
  } else if (!is.null(country) & is.null(year)) {
    browsable(div(style = "max-height:500px; overflow-y:auto;", 
      measles |>
      dplyr::filter(Country == country) |>
      dplyr::select(Country, Year, `Nominal GDP per Capita`, 
                    `MCV1 Coverage`, `Total Measles per Million`) |>
        arrange(desc(`Total Measles per Million`)) |>
        rename("Nominal GDP per Capita ($ USD)" = `Nominal GDP per Capita`) |>
      gt() |>
      fmt_number(columns = c(`Nominal GDP per Capita ($ USD)`), decimals = 3) |>
      fmt(columns = `MCV1 Coverage`,
          fns = function(x) paste0(x, "%")) |>
      tab_style(style = cell_text(weight = "bold"),
                locations = cells_column_labels(everything()))))
    
  } else {
    stop("Invalid combination of inputs. Provide either country (+ optional year) or region + year.")}}

mcv_gdp(region = "western pacific", year = 2016)
```



# Romania Measles Data

## Single Column

### Romania: representative case
Based on our Global Annual Measles Cases Rates per Million table, along with the heatmap, we noticed that Europe 2024 yielded the most recent outbreak of measles cases (last 5 years). Upon examining this region and year further, we noticed that Romania was roughly representative of Europe in terms of population with 19.07 million people and also possessed the largest total measles cases and third largest average cases in Europe 2024. A deeper introspection into potential socioeconomic factors in Romania during 2024 could help us understand the spike in cases.

Despite decreasing in Measles Coverage Vaccine Rate, socioeconomic factors such as GDP, homelessness, etc. do not appear to decrease. However, we do see a similar decreasing trend for MCV1 Coverage in primary school completion rate. We hypothesize decreasing education rate might be associated with vaccination rate and thus increased measles cases.

### Average MCV1 Coverage Compared to 

```{r}
#| title: Average MCV1 Coverage for 1 Year Olds Across Time and Region (WHO) Compared to Primary School Completion Rate
# make plot

measles |>
  filter(Country == "Romania") |>
  filter(!is.na(`MCV1 Coverage`)) |>
left_join(romania, by = join_by("Year" == "name")) |>
  rename("Primary School Completion Rate (%)" = value,
         "MCV1 Coverage (%)" = `MCV1 Coverage`) |>
  pivot_longer(cols = c(`MCV1 Coverage (%)`, `Primary School Completion Rate (%)`), names_to = "series",
    values_to = "value") |>
  mutate(series = as.factor(series)) |>
  ggplot(aes(x = Year, y = value, color = "orange")) + geom_point() + geom_line() + labs(
    title = "Romania: {closest_state}",
    x = "Year",
    y = "",
    color = ""
  ) +
  theme_minimal() + theme(panel.grid.minor = element_blank()) + scale_x_continuous(breaks = c(2015, 2020, 2024)) + scale_y_continuous(labels = scales::percent_format(scale = 1), limits = c(70, 100)) + transition_states(series, transition_length = 2,
                                                               state_length = 1)
```
